diffEnrich: An \proglang{R} Package to Compare Functional Enrichment Between Two Experimentally-derived Groups of Genes by Connecting to the KEGG REST API

### 2019-12-05 jss-article-rnw-diffEnrich/article.tex
incorp supp info into text body and remove supp section

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Kegg Release: Release\_92.0+\_11-23\_Nov\_19}}\end{kframe}
\end{knitrout}

@@

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# ncbi\_to\_kegg2019-11-22Release\_92.0+\_11-23\_Nov\_19.txt}}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# kegg\_to\_pathway2019-11-22Release\_92.0+\_11-23\_Nov\_19.txt}}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# pathway\_to\_species2019-11-22Release\_92.0+\_11-23\_Nov\_19.txt}}

@@

By default, p-values from the Fisher’s Exact test are adjusted for multiple
comparisons with a False Discovery Rate (FDR) (Benjamini:1995), however users
have the option of choosing any type of multiple testing correction supported
by \code{p.adjust}. In addition to the unadjusted p-value and FDR, pathEnrich
will calculate for each KEGG pathway, its fold enrichment defined as the ratio
of number of genes observed from the gene list annotated to the KEGG pathway to
the expected number of genes from the gene list to be annotated to the KEGG
pathway by chance. An example of the first 6 results generated by \code{pathEnrich}
are in Supplementary Table 1. For a detailed description of the variables in this table see Table 2.

@@

S3 generic functions for \code{print} and \code{summary} are provided.
The \code{print} function prints the results table as a \code{tibble}, and the
\code{summary} function returns the number of pathways that reached statistical
significance as well as their descriptions, the number of genes used from the
KEGG data base, the KEGG species, and the method used for multiple testing
correction. See supplementary information for examples.

@@

@@

@@

%% -----Supplementary Information------------------------------------
\section[Supplementary Information]{Supplementary Information} \label{sec:supinfo}

\begin{table}[H]
  \centering
  \caption{Supplementary Table 1. First 6 results generated by pathEnrich.}
  \begin{adjustbox}{width=\textwidth, angle=90}
  \fontsize{90}{125}\selectfont
  \begin{tabular}{rllrrrrrrrr}
  \hline
 & KEGG\_PATHWAY\_ID & KEGG\_PATHWAY\_description & KEGG\_PATHWAY\_cnt & KEGG\_PATHWAY\_in\_list & KEGG\_DATABASE\_cnt & KEG\_DATABASE\_in\_list & expected & enrich\_p & p\_adj & fold\_enrichment \\
  \hline
95 & rno04530 & Tight junction - Rattus norvegicus (rat) & 170 & 19.00 & 8856 & 295 & 5.66 & 0.00 & 0.00 & 3.36 \\
  172 & rno05135 & Yersinia infection - Rattus norvegicus (rat) & 128 & 16.00 & 8856 & 295 & 4.26 & 0.00 & 0.00 & 3.75 \\
  194 & rno05210 & Colorectal cancer - Rattus norvegicus (rat) &  88 & 12.00 & 8856 & 295 & 2.93 & 0.00 & 0.00 & 4.09 \\
  212 & rno05231 & Choline metabolism in cancer - Rattus norvegicus (rat) &  99 & 12.00 & 8856 & 295 & 3.30 & 0.00 & 0.00 & 3.64 \\
  197 & rno05213 & Endometrial cancer - Rattus norvegicus (rat) &  58 & 9.00 & 8856 & 295 & 1.93 & 0.00 & 0.00 & 4.66 \\
  66 & rno04144 & Endocytosis - Rattus norvegicus (rat) & 275 & 22.00 & 8856 & 295 & 9.16 & 0.00 & 0.00 & 2.40 \\
   \hline
  \end{tabular}
\end{adjustbox}
\end{table}

\begin{table}[H]
\centering
  \caption{Supplementary Table 2. First 6 results generated by diffEnrich}
  \begin{adjustbox}{width=\textwidth, angle=90}
  \fontsize{50}{80}\selectfont
\begin{tabular}{rllrrrrrrrrrrrrrrrrr}
  \hline
 & KEGG\_PATHWAY\_ID & KEGG\_PATHWAY\_description & KEGG\_PATHWAY\_cnt & KEGG\_DATABASE\_cnt & KEGG\_PATHWAY\_in\_list1 & KEGG\_DATABASE\_in\_list1 & expected\_list1 & enrich\_p\_list1 & p\_adj\_list1 & fold\_enrichment\_list1 & KEGG\_PATHWAY\_in\_list2 & KEGG\_DATABASE\_in\_list2 & expected\_list2 & enrich\_p\_list2 & p\_adj\_list2 & fold\_enrichment\_list2 & odd\_ratio & diff\_enrich\_p & diff\_enrich\_adjusted \\
  \hline
rno04530 & rno04530 & Tight junction - Rattus norvegicus (rat) & 170 & 8856 & 19.00 & 295 & 5.66 & 0.00 & 0.00 & 3.36 & 131.00 & 5308 & 101.89 & 0.00 & 0.00 & 1.29 & 0.37 & 0.00 & 0.00 \\
  rno05135 & rno05135 & Yersinia infection - Rattus norvegicus (rat) & 128 & 8856 & 16.00 & 295 & 4.26 & 0.00 & 0.00 & 3.75 & 105.00 & 5308 & 76.72 & 0.00 & 0.00 & 1.37 & 0.35 & 0.00 & 0.00 \\
  rno05210 & rno05210 & Colorectal cancer - Rattus norvegicus (rat) &  88 & 8856 & 12.00 & 295 & 2.93 & 0.00 & 0.00 & 4.09 & 81.00 & 5308 & 52.74 & 0.00 & 0.00 & 1.54 & 0.37 & 0.00 & 0.00 \\
  rno05213 & rno05213 & Endometrial cancer - Rattus norvegicus (rat) &  58 & 8856 & 9.00 & 295 & 1.93 & 0.00 & 0.00 & 4.66 & 55.00 & 5308 & 34.76 & 0.00 & 0.00 & 1.58 & 0.33 & 0.01 & 0.01 \\
  rno04660 & rno04660 & T cell receptor signaling pathway - Rattus norvegicus (rat) & 106 & 8856 & 11.00 & 295 & 3.53 & 0.00 & 0.01 & 3.12 & 79.00 & 5308 & 63.53 & 0.00 & 0.00 & 1.24 & 0.39 & 0.01 & 0.01 \\
  rno04657 & rno04657 & IL-17 signaling pathway - Rattus norvegicus (rat) &  95 & 8856 & 9.00 & 295 & 3.16 & 0.00 & 0.03 & 2.84 & 59.00 & 5308 & 56.94 & 0.37 & 0.46 & 1.04 & 0.36 & 0.01 & 0.01 \\
   \hline
\end{tabular}
\end{adjustbox}
\end{table}

Example of S3 functions
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{summary}\hlstd{(list1_pe)}
\end{alltt}
\begin{verbatim}
## 219 KEGG pathways were tested. 
##  Only KEGG pathways that contained at least 2 genes from gene_list were tested. 
##  KEGG pathway species: Rattus norvegicus (rat)
##  8856 genes from gene_list were in the KEGG data pull. 
##  p-value adjustment method: BH
##  36 pathways reached statistical significance after multiple testing correction at a cutoff of 0.05. 
##  
## Significant pathways: 
##  Tight junction
## Yersinia infection
## Colorectal cancer
## Choline metabolism in cancer
## Endometrial cancer
## Endocytosis
## Neurotrophin signaling pathway
## Thermogenesis
## Oocyte meiosis
## VEGF signaling pathway
## Thyroid hormone signaling pathway
## Hippo signaling pathway
## T cell receptor signaling pathway
## Apoptosis
## Hepatocellular carcinoma
## MAPK signaling pathway
## Focal adhesion
## Salmonella infection
## Non-alcoholic fatty liver disease (NAFLD)
## ErbB signaling pathway
## Sphingolipid signaling pathway
## Pancreatic cancer
## Progesterone-mediated oocyte maturation
## Alzheimer disease
## Endocrine resistance
## Adrenergic signaling in cardiomyocytes
## IL-17 signaling pathway
## Chronic myeloid leukemia
## Dopaminergic synapse
## Prostate cancer
## EGFR tyrosine kinase inhibitor resistance
## Hepatitis C
## Ras signaling pathway
## Acute myeloid leukemia
## Insulin signaling pathway
## Fc epsilon RI signaling pathway
\end{verbatim}
\end{kframe}
\end{knitrout}

>>>>

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Kegg Release: Release\_92.0+\_12-05\_Dec\_19}}\end{kframe}
\end{knitrout}

@@

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# ncbi\_to\_kegg2019-12-05Release\_92.0+\_12-05\_Dec\_19.txt}}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# kegg\_to\_pathway2019-12-05Release\_92.0+\_12-05\_Dec\_19.txt}}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# pathway\_to\_species2019-12-05Release\_92.0+\_12-05\_Dec\_19.txt}}

@@

By default, p-values from the Fisher’s Exact test are adjusted for multiple
comparisons with a False Discovery Rate (FDR) (Benjamini:1995), however users
have the option of choosing any type of multiple testing correction supported
by \code{p.adjust}. In addition to the unadjusted p-value and FDR, pathEnrich
will calculate for each KEGG pathway, its fold enrichment defined as the ratio
of number of genes observed from the gene list annotated to the KEGG pathway to
the expected number of genes from the gene list to be annotated to the KEGG
pathway by chance. An example of the first 6 results generated by \code{pathEnrich}
are in displayed below. For a detailed description of the variables in this table see Table 2.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{head}\hlstd{(list1_pe}\hlopt{$}\hlstd{enrich_table)}
\end{alltt}
\begin{verbatim}
##     KEGG_PATHWAY_ID
## 95         rno04530
## 172        rno05135
## 194        rno05210
## 212        rno05231
## 197        rno05213
## 66         rno04144
##                                   KEGG_PATHWAY_description
## 95                Tight junction - Rattus norvegicus (rat)
## 172           Yersinia infection - Rattus norvegicus (rat)
## 194            Colorectal cancer - Rattus norvegicus (rat)
## 212 Choline metabolism in cancer - Rattus norvegicus (rat)
## 197           Endometrial cancer - Rattus norvegicus (rat)
## 66                   Endocytosis - Rattus norvegicus (rat)
##     KEGG_PATHWAY_cnt KEGG_PATHWAY_in_list KEGG_DATABASE_cnt
## 95               170                   19              8856
## 172              128                   16              8856
## 194               88                   12              8856
## 212               99                   12              8856
## 197               58                    9              8856
## 66               275                   22              8856
##     KEG_DATABASE_in_list expected     enrich_p        p_adj
## 95                   295 5.662827 3.485551e-06 0.0005387284
## 172                  295 4.263776 4.919894e-06 0.0005387284
## 194                  295 2.931346 3.179635e-05 0.0023211334
## 212                  295 3.297764 1.032192e-04 0.0044701134
## 197                  295 1.932023 1.132216e-04 0.0044701134
## 66                   295 9.160456 1.224689e-04 0.0044701134
##     fold_enrichment
## 95         3.355214
## 172        3.752542
## 194        4.093683
## 212        3.638829
## 197        4.658328
## 66         2.401627
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

S3 generic functions for \code{print} and \code{summary} are provided.
The \code{print} function prints the results table as a \code{tibble}, and the
\code{summary} function returns the number of pathways that reached statistical
significance as well as their descriptions, the number of genes used from the
KEGG data base, the KEGG species, and the method used for multiple testing
correction.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{summary}\hlstd{(list1_pe)}
\end{alltt}
\begin{verbatim}
## 219 KEGG pathways were tested. 
##  Only KEGG pathways that contained at least 2 genes from gene_list were tested. 
##  KEGG pathway species: Rattus norvegicus (rat)
##  8856 genes from gene_list were in the KEGG data pull. 
##  p-value adjustment method: BH
##  36 pathways reached statistical significance after multiple testing correction at a cutoff of 0.05. 
##  
## Significant pathways: 
##  Tight junction
## Yersinia infection
## Colorectal cancer
## Choline metabolism in cancer
## Endometrial cancer
## Endocytosis
## Neurotrophin signaling pathway
## Thermogenesis
## Oocyte meiosis
## VEGF signaling pathway
## Thyroid hormone signaling pathway
## Hippo signaling pathway
## T cell receptor signaling pathway
## Apoptosis
## Hepatocellular carcinoma
## MAPK signaling pathway
## Focal adhesion
## Salmonella infection
## Non-alcoholic fatty liver disease (NAFLD)
## ErbB signaling pathway
## Sphingolipid signaling pathway
## Pancreatic cancer
## Progesterone-mediated oocyte maturation
## Alzheimer disease
## Endocrine resistance
## Adrenergic signaling in cardiomyocytes
## IL-17 signaling pathway
## Chronic myeloid leukemia
## Dopaminergic synapse
## Prostate cancer
## EGFR tyrosine kinase inhibitor resistance
## Hepatitis C
## Ras signaling pathway
## Acute myeloid leukemia
## Insulin signaling pathway
## Fc epsilon RI signaling pathway
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{head}\hlstd{(diff_enrich}\hlopt{$}\hlstd{de_table)}
\end{alltt}
\begin{verbatim}
##          KEGG_PATHWAY_ID
## rno04530        rno04530
## rno05135        rno05135
## rno05210        rno05210
## rno05213        rno05213
## rno04660        rno04660
## rno04657        rno04657
##                                             KEGG_PATHWAY_description
## rno04530                    Tight junction - Rattus norvegicus (rat)
## rno05135                Yersinia infection - Rattus norvegicus (rat)
## rno05210                 Colorectal cancer - Rattus norvegicus (rat)
## rno05213                Endometrial cancer - Rattus norvegicus (rat)
## rno04660 T cell receptor signaling pathway - Rattus norvegicus (rat)
## rno04657           IL-17 signaling pathway - Rattus norvegicus (rat)
##          KEGG_PATHWAY_cnt KEGG_DATABASE_cnt KEGG_PATHWAY_in_list1
## rno04530              170              8856                    19
## rno05135              128              8856                    16
## rno05210               88              8856                    12
## rno05213               58              8856                     9
## rno04660              106              8856                    11
## rno04657               95              8856                     9
##          KEGG_DATABASE_in_list1 expected_list1 enrich_p_list1
## rno04530                    295       5.662827   3.485551e-06
## rno05135                    295       4.263776   4.919894e-06
## rno05210                    295       2.931346   3.179635e-05
## rno05213                    295       1.932023   1.132216e-04
## rno04660                    295       3.530939   7.743416e-04
## rno04657                    295       3.164521   4.270915e-03
##           p_adj_list1 fold_enrichment_list1 KEGG_PATHWAY_in_list2
## rno04530 0.0005387284              3.355214                   131
## rno05135 0.0005387284              3.752542                   105
## rno05210 0.0023211334              4.093683                    81
## rno05213 0.0044701134              4.658328                    55
## rno04660 0.0129121321              3.115318                    79
## rno04657 0.0346418644              2.844032                    59
##          KEGG_DATABASE_in_list2 expected_list2 enrich_p_list2
## rno04530                   5308      101.89250   1.459992e-06
## rno05135                   5308       76.71906   5.537739e-08
## rno05210                   5308       52.74435   9.806872e-12
## rno05213                   5308       34.76332   1.160182e-09
## rno04660                   5308       63.53297   1.107477e-03
## rno04657                   5308       56.93993   3.737223e-01
##           p_adj_list2 fold_enrichment_list2 odd_ratio diff_enrich_p
## rno04530 5.568342e-06              1.285669 0.3676651  0.0002935876
## rno05135 2.594826e-07              1.368630 0.3520039  0.0005434729
## rno05210 1.237175e-10              1.535709 0.3655602  0.0032572306
## rno05213 8.849757e-09              1.582127 0.3328275  0.0058774642
## rno04660 2.256227e-03              1.243449 0.3901694  0.0072047984
## rno04657 4.591045e-01              1.036180 0.3572935  0.0087537607
##          diff_enrich_adjusted
## rno04530         0.0002935876
## rno05135         0.0005434729
## rno05210         0.0032572306
## rno05213         0.0058774642
## rno04660         0.0072047984
## rno04657         0.0087537607
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{summary}\hlstd{(diff_enrich)}
\end{alltt}
\begin{verbatim}
## 219 KEGG pathways were shared between gene lists and were tested. 
##  KEGG pathway species: Rattus norvegicus (rat)
##  8856 genes from gene_list were in the KEGG data pull. 
##  p-value adjustment method: none
##  19 pathways reached statistical significance after multiple testing correction at a cutoff of 0.05. 
##  
## Significant pathways: 
##  Tight junction
## Yersinia infection
## Colorectal cancer
## Endometrial cancer
## T cell receptor signaling pathway
## IL-17 signaling pathway
## Salmonella infection
## Choline metabolism in cancer
## Endocytosis
## VEGF signaling pathway
## Oocyte meiosis
## Thermogenesis
## Hippo signaling pathway
## Neurotrophin signaling pathway
## Apoptosis
## Hepatocellular carcinoma
## Fc epsilon RI signaling pathway
## Thyroid hormone signaling pathway
## Alzheimer disease
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

%% -----Supplementary Information------------------------------------
% \section[Supplementary Information]{Supplementary Information} \label{sec:supinfo}

% \begin{table}[H]
%   \centering
%   \caption{Supplementary Table 1. First 6 results generated by pathEnrich.}
%   \begin{adjustbox}{width=\textwidth, angle=90}
%   \fontsize{90}{125}\selectfont
%   \begin{tabular}{rllrrrrrrrr}
%   \hline
%  & KEGG\_PATHWAY\_ID & KEGG\_PATHWAY\_description & KEGG\_PATHWAY\_cnt & KEGG\_PATHWAY\_in\_list & KEGG\_DATABASE\_cnt & KEG\_DATABASE\_in\_list & expected & enrich\_p & p\_adj & fold\_enrichment \\
%   \hline
% 95 & rno04530 & Tight junction - Rattus norvegicus (rat) & 170 & 19.00 & 8856 & 295 & 5.66 & 0.00 & 0.00 & 3.36 \\
%   172 & rno05135 & Yersinia infection - Rattus norvegicus (rat) & 128 & 16.00 & 8856 & 295 & 4.26 & 0.00 & 0.00 & 3.75 \\
%   194 & rno05210 & Colorectal cancer - Rattus norvegicus (rat) &  88 & 12.00 & 8856 & 295 & 2.93 & 0.00 & 0.00 & 4.09 \\
%   212 & rno05231 & Choline metabolism in cancer - Rattus norvegicus (rat) &  99 & 12.00 & 8856 & 295 & 3.30 & 0.00 & 0.00 & 3.64 \\
%   197 & rno05213 & Endometrial cancer - Rattus norvegicus (rat) &  58 & 9.00 & 8856 & 295 & 1.93 & 0.00 & 0.00 & 4.66 \\
%   66 & rno04144 & Endocytosis - Rattus norvegicus (rat) & 275 & 22.00 & 8856 & 295 & 9.16 & 0.00 & 0.00 & 2.40 \\
%    \hline
%   \end{tabular}
% \end{adjustbox}
% \end{table}

% \begin{table}[H]
% \centering
%   \caption{Supplementary Table 2. First 6 results generated by diffEnrich}
%   \begin{adjustbox}{width=\textwidth, angle=90}
%   \fontsize{50}{80}\selectfont
% \begin{tabular}{rllrrrrrrrrrrrrrrrrr}
%   \hline
%  & KEGG\_PATHWAY\_ID & KEGG\_PATHWAY\_description & KEGG\_PATHWAY\_cnt & KEGG\_DATABASE\_cnt & KEGG\_PATHWAY\_in\_list1 & KEGG\_DATABASE\_in\_list1 & expected\_list1 & enrich\_p\_list1 & p\_adj\_list1 & fold\_enrichment\_list1 & KEGG\_PATHWAY\_in\_list2 & KEGG\_DATABASE\_in\_list2 & expected\_list2 & enrich\_p\_list2 & p\_adj\_list2 & fold\_enrichment\_list2 & odd\_ratio & diff\_enrich\_p & diff\_enrich\_adjusted \\
%   \hline
% rno04530 & rno04530 & Tight junction - Rattus norvegicus (rat) & 170 & 8856 & 19.00 & 295 & 5.66 & 0.00 & 0.00 & 3.36 & 131.00 & 5308 & 101.89 & 0.00 & 0.00 & 1.29 & 0.37 & 0.00 & 0.00 \\
%   rno05135 & rno05135 & Yersinia infection - Rattus norvegicus (rat) & 128 & 8856 & 16.00 & 295 & 4.26 & 0.00 & 0.00 & 3.75 & 105.00 & 5308 & 76.72 & 0.00 & 0.00 & 1.37 & 0.35 & 0.00 & 0.00 \\
%   rno05210 & rno05210 & Colorectal cancer - Rattus norvegicus (rat) &  88 & 8856 & 12.00 & 295 & 2.93 & 0.00 & 0.00 & 4.09 & 81.00 & 5308 & 52.74 & 0.00 & 0.00 & 1.54 & 0.37 & 0.00 & 0.00 \\
%   rno05213 & rno05213 & Endometrial cancer - Rattus norvegicus (rat) &  58 & 8856 & 9.00 & 295 & 1.93 & 0.00 & 0.00 & 4.66 & 55.00 & 5308 & 34.76 & 0.00 & 0.00 & 1.58 & 0.33 & 0.01 & 0.01 \\
%   rno04660 & rno04660 & T cell receptor signaling pathway - Rattus norvegicus (rat) & 106 & 8856 & 11.00 & 295 & 3.53 & 0.00 & 0.01 & 3.12 & 79.00 & 5308 & 63.53 & 0.00 & 0.00 & 1.24 & 0.39 & 0.01 & 0.01 \\
%   rno04657 & rno04657 & IL-17 signaling pathway - Rattus norvegicus (rat) &  95 & 8856 & 9.00 & 295 & 3.16 & 0.00 & 0.03 & 2.84 & 59.00 & 5308 & 56.94 & 0.37 & 0.46 & 1.04 & 0.36 & 0.01 & 0.01 \\
%    \hline
% \end{tabular}
% \end{adjustbox}
% \end{table}
